---
title: "Downloading ERA5 data for use with `microclima`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

ERA5 climate data can be downloaded from the climate data store (CDS). The 
following describes how to access the data using R:

1) Register [here](https://cds.climate.copernicus.eu/user/register). 

2) Then, navigate [here](https://cds.climate.copernicus.eu/), login, and click on 
your name (in the top right hand corner) to access your profile. Here you will 
find your user ID (UID) and API key, both which are required for you to remotely 
download data from the CDS. Make a note of these.  

3) You will also need to accept the Ts&Cs (whilst logged in) [here](https://cds.climate.copernicus.eu/cdsapp/#!/terms/licence-to-use-copernicus-products).

The following packages are required:
```{r packages, warning = FALSE, message = FALSE}
library(dplyr)
library(ecmwfr)
library(keyring)
library(lubridate)
library(tidync)
library(microclima) # remotes::install_github("ilyamaclean/microclima")
library(NicheMapR) # remotes::install_github("mrke/NicheMapR")
```

```{r funs, include = FALSE}
#############
##FUNCTIONS##
#############

files_source <- list.files(here::here("R/"), full.names = T)
sapply(files_source,source)
```

#### Set user credentials for API access

```{r creds, eval = FALSE}
# assign your credentials
uid <- "*****"
cds_api_key <- "********-****-****-****-************"

ecmwfr::wf_set_key(user = uid,
                   key = cds_api_key,
                   service = "cds")
```

## Usage

#### Building a request

The first step is to decide on the spatial and temporal extents of the data retrieved from the CDS.
There is a fair sized overhead in submitting a request (and consequently waiting in a queue) so it 
is sensible to execute as few requests as possible. To this end, it is recommended to ensure the 
spatial extent covers multiple areas of interest (within reason) and that the temporal extent covers
multiple time periods of interest.

Requests are submitted as whole months, and also they will be split by year, so that they are not too big to handle. The splitting of requests which overlap year boundaries is dealt with automatically. 

Once these parameters are decided, one can begin to build the request(s) as follows:

```{r build request}

# bounding coordinates (in WGS84 / EPSG:4326)
xmn <- -4
xmx <- -2
ymn <- 49
ymx <- 51

# temporal extent
st_time <- lubridate::ymd("2010:02:26")
en_time <- lubridate::ymd("2010:03:01")

# filename and location for downloaded .nc files
file_prefix <- "era5_test"
op <- "C:/Sandbox"

# build a request (covering multiple years)
req <- build_era5_request(xmin = xmn, xmax = xmx, 
                          ymin = ymn, ymax = ymx,
                          start_time = st_time,
                          end_time = en_time,
                          outfile_name = file_prefix)
```

Requests are stored in list format. Each request (divided by year) are stored as list
objects within the master list:

```{r list_view}
str(req)
```

#### Obtaining data with a request

The next step is to execute the requests, by sending them to the CDS. The `request_era5` function will deal with lists containing multiple requests (i.e. those created by `build_era5_request` with temporal extents spanning multiple years). Executing `request_era5` will download .nc files for each year to the location defined by `out_path`. The filenames will have a prefix defined by `outfile_name` in `build_era5_request`:

```{r send_request, eval = FALSE}
request_era5(request = req, uid = uid, out_path = op)
```

#### Processing .nc files

Once all .nc files are downloaded, the next step is to extract climate variables conforming to desired spatial and temporal extents from them. This can be done for a single point in space with `extract_clim`. The function outputs a dataframe where each row is a single observation in time and space and each column is a climatic variable. Note that this function uses the same start and end times defined prior to building the request. Furthermore, the function (by default) uses the land-sea mask data (automatically downloaded alongside climate variables with `request_era5`), to filter out points that do not meet a land-sea threshold. 0 represents complete sea, and 1 complete land. By default, the threshold is liberally set to 0.05. Given that `runauto` in the `microclima` package only accepts date ranges within single years, you will need to create a data frame for each yearly block. For example, if your period of interest spans multiple years (and you have used the previous code to download multiple .nc files, run this next block of code for each year):

```{r process_nc, eval = FALSE}
# list the path of the .nc file for a given year
my_nc <- "C:/Sandbox/era5_test_2010.nc"

# for a single point (make sure it's within the bounds of your .nc file)
long <- -3.654600
lat <- 50.640369

# gather all hourly variables
point_out <- extract_clim(nc = my_nc, x = long, y = lat,
                            start_time = st_time,  end_time = en_time)
head(point_out)

# gather daily precipitation
point_out_precip <- extract_precip(nc = my_nc, x = long, y = lat,
                                          start_time = st_time,  
                                          end_time = en_time)
```

The dataframe created by `extract_clim` and vector created by `extract_precip` are ready to be used as inputs to the `runauto` function from the `microclima` package:

```{r runauto_example, eval = FALSE}
# create a 200 x 200 30 m spatial resoltuion DEM for location
r <- microclima::get_dem(lat = lat, long = long, resolution = 30)

# change date format to fit runauto requirements
temps <- microclima::runauto(r = r, dstart = "26/02/2010",dfinish = "01/03/2010", 
                             hgt = 0.1, l = NA, x = NA, 
                             habitat = "Barren or sparsely vegetated",
                             hourlydata = point_out, 
                             dailyprecip = point_out_precip, 
                             plot.progress= FALSE)
```

#### Acknowledgements

Thank you to Koen Hufkens, creator of the `ecwmfr` package for making the aquisition of ECWMF data through R possible. 
